---
title: "Lending Club Data Analysis"
author: "Chengdong Liang"
date: "November 2, 2016"
output:
  html_document: default
  word_document: default
---

```{r change wd, include=FALSE}
setwd("C:/Users/Chengdong Liang/Desktop/Lending Club Data")
```
```{r package library, include=FALSE}
library(choroplethr) #Simplify the Creation of Choropleth Maps in R
library(choroplethrMaps) #Contains maps used bu the choroplethr package
library(DescTools) #A collection of miscellaneous basic statistic funtions and convenience wrappers for efficiently describing data.
library(dplyr) #A fast, consistent tool for working with data frame like objects, both in memory and out of memory.
library(ggplot2) #An implementation of the grammar of graphics
library(readr) #Read tabular data
```
```{r import data, include=FALSE}
options(scipen=999) #Force R not to use exponential notation
loanbook<- read_csv("lending-club-loan-data/loan.csv") #read.csv is much more slower

#load the state names
data(state.regions)

#merge the loan book with the state names
loanbook <- merge(loanbook,state.regions,by.x="addr_state",by.y="abb")
```

## Data overview

After loading the entire data set I would like to check its structure and summarise it.

```{r dim and colnames, echo=TRUE}
# print dimentions
dim(loanbook)

#print column names
colnames(loanbook)
```

I am not printing these results for conciseness:

```{r str and summary, eval=FALSE} 
str(loanbook) #eval=FALSE, to display R code without evaluating it.
summary(loanbook)
```

## Missing variables

I've also noticed that several key fields described in the data dictionary were missing from the loan book. In addition, there are also several key fields appear in the loan book are missing from the data dictionary. There might be some inconsistent:

```{r, echo=TRUE}
library(readxl) #Read Excel Files

dataDictionary<- read_excel("lending-club-loan-data/LCDataDictionary.xlsx")

# fields available in the data dictionary
dd_names <- as.character(na.omit(dataDictionary$LoanStatNew))

# fileds available in the loan book
loanbook_names <- names(loanbook)

# show the fields described in data dictionary but not in the loan book
setdiff(dd_names, loanbook_names) #

# show the fields described in the loan book but not in data dictionary
setdiff(loanbook_names, dd_names)
```
## Loan amounts

I am interested in several variables and I plotted them using DescTools package.
Here I created the density plot, box plot and empirical distribution function plot.
```{r, echo=TRUE}
Desc(loanbook$loan_amnt, main="Loan amount distribution", plotit=TRUE)
```

Here is how the loan book was growing:
```{r, echo=TRUE}
loanbook$issue_d <-as.Date(gsub("^","01-",loanbook$issue_d), format="%d-%b-%Y")

amnt_df <- loanbook %>% #Learn from it
  select(issue_d, loan_amnt) %>%
  group_by(issue_d) %>%
  summarise(Amount= sum(loan_amnt))

ts_amnt <- ggplot(amnt_df, aes(x= issue_d, y=Amount))

ts_amnt + geom_line() + xlab("Date issued")
```

## Loan statues

Terms interpretation:

Charged off: Loan for which there is no longer a reasonable expectation of further payments. Generally, Charge Off occurs no later than 30 days after the Default status is reached. Upon Charge Off, the remaining principal balance of the Note is deducted from the account balance. Charge Off typically occurs when a loan is 150 days past due (i.e. 30 days after the Default status is reached) and there is no reasonable expectation of sufficient payment to prevent the charge off. 

Default: Loan has not been current for 121 days or more.
The loanbook consists of loans in various statuses so I started with exploring their frequency. In general, a Note enters Default status when it is 121+ days past due.

In Grace Period: Loan is past due but within the 15-day grace period.

Current: Loan is up to date on all outstanding payments. 
```{r statuses frequency}
Desc(loanbook$loan_status, plotit=TRUE) #categorical variable, same code with different plot
```

Then I checked the distribution of loan amounts by status.
```{r loan amounts by status, echo=TRUE}
box_status <- ggplot(loanbook, aes(loan_status, loan_amnt))
box_status + geom_boxplot(aes(fill=loan_status)) + 
  theme(axis.text.x= element_blank()) +
  labs(list(
    title= "Loan amount by status",
    x= "Status",
    y="Amount")) 
# box_status gives basic outline of the plot. Namely x axis and y axis
# geom_boxplot(aes(fill=loan_status)) filled the boxplot
# theme(axis.text.x= element_blank()) eliminate axis.text.x 
# labs(list(title= "Loan amount by status", x= "Status", y="Amount")) given titles
```

Here is how the value of loans of different grades was changing over time
```{r loan amount with grades changing by time, echo= TRUE}
amnt_df_grade <- loanbook %>%
  select(issue_d, loan_amnt, grade) %>%
  group_by(issue_d, grade ) %>%
  summarise(Amount= sum(loan_amnt))

ts_amnt_grade <- ggplot(amnt_df_grade, 
                        aes(x= issue_d, y= Amount))

ts_amnt_grade + geom_area(aes(fill= grade)) +xlab("Date issued")
```

## Maps

In oder to create maps of loans the author initially joined the loan book with names of states and then the author createda data frame required by choroplethr.

### Loans by value
```{r loans by value, echo= TRUE}
state_by_value <-
  loanbook %>% group_by(region) %>%
  summarise(value= sum(loan_amnt, na.rm= TRUE))

state_choropleth(state_by_value, title = "Value by State")
```

### Loans by volume
```{r, echo= TRUE}
state_by_volume <-
  loanbook %>%
  group_by(region) %>%
  summarise(value=n())

state_choropleth(state_by_volume, title= "Volume by state")
```

### Loan reasons

### What's the reason for taking a loan with LengingClub?

```{r loan reason}
Desc(loanbook$purpose, main= "Loan purposes", plotit= TRUE)
Desc(loanbook$title, main= "Loan titles", plotit= TRUE)
```

## Word cloud

Word cloud gives a good overview of the loan titles (gives by borrowers).
This information should explain what kind of loans are bing funded by LendingClub.

In order to create a word cloud, First the author loaded the necessary libraries, 
then he preprocessed the 'title' column by removing punctuation and transforming it to lower case.
This analysis was run on the first 10000 rows to speed up the process.

```{r word cloud, echo=TRUE, warning=FALSE, message=FALSE}
library(tm)
library(RColorBrewer)
library(wordcloud)

loan_descriptions.corpus <- Corpus(DataframeSource(data.frame(head(loanbook[,23],n=10000))))
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removePunctuation)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, content_transformer(tolower))

wordcloud(loan_descriptions.corpus, 
          max.words = 100,
          random.order= FALSE, #plotted in decreasing frequency
          rot.per = 0.30, #proportion words with 90 degree rotation
          use.r.layout = FALSE, #C++ code is used for collision detection
          colors = brewer.pal(8, "Paired"))

```
## Loan grades

Here is the overview of the occurrence of loans of different grades:
```{r loan grades, echo=TRUE}
Desc(loanbook$grade, main= "Loan grades", plotit=TRUE)
```

The last step (so far) was checking whether the interest rates are dependent on the loan grade.
```{r interest&grade}
Desc(int_rate~grade, loanbook, digits= 1, main= "Interest rate by grade", plotit= TRUE)
```

Unsurprisingly, the higher the grade (more risky loan), the higher the interest rates.

## Default prediction

There are several statuses which indicate that loans are not perfroming well.
The author put them into two groups.

```{r default prediction}
#'bad' statuses
bad_indicators <- c("Charged off",
                    "Default",
                    "Does not meet the credit policy. Status:Charged Off",
                    "In Grace Period", 
                    "Default Receiver",
                     "Late (16-30 days)",
                      "Late (31-120 days)")

# assign certain statuses to a 'bad' ('0') group
loanbook$is_bad <- ifelse(loanbook$loan_status %in% bad_indicators, 0,                                           ifelse(loanbook$loan_status=="", NA, 1)
                          ) 
```

Then the author checks wheher there is a difference between the 'bad' abd 'good' loans for numeric variables.

```{r bad and good loans}
#figure out which columns are numeric so that we can look at the distribution
numeric_cols <- sapply(loanbook, is.numeric)

# turn the data into long format
library(reshape2)
loanbook.lng <- melt(loanbook[,numeric_cols], id="is_bad")

#plot the distribution for 'bad' and 'good' for each numeric variable
p <- ggplot(aes(x= value, group= is_bad, colour= factor(is_bad)),
            data= loanbook.lng)

#create the plot to check if there are any good variables that can be used in predictive models
p + geom_density() + 
  facet_wrap(~variable, scales="free")
```

The plot above can be the basis for chosing variables for predictive models (note: scales need fixing).

It seems like Annual Income and Interest Rate are two variables that can be good predictors of how loans will behave
We can have a closer look at particular cases:

```{r}
library(DT)

loanbook %>%
  filter(is_bad=='0') %>%
  select(annual_inc, int_rate, loan_status) %>%
  datatable(., options= list(pageLength= 10))
```


